---
title: "gapminder-wrangle.Rmd"
author: "Gretel Waugh"
date: "30/03/2020"
output: html_document
---
#Here we go...

We've moved onto Chapter 6, where I'll be learning to wrangle data using dplyr. Tidyverse is already installed, but I'll need to reload because we restarted R:

```{r tidyverse}

library(tidyverse)
```

Lights on!

Now we're ready to read in the data, which is from GitHub. 

```{r ohidata}
gapminder <- read.csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")
```

Strangely, readr wouldn't work for me. I used read.csv instead.
Now we can inspect the data:

```{r inspect}
head(gapminder)
tail(gapminder, 10) #specifying a number will show that many lines of the data
```

This is a new command for me: str(). It shows the structure of an object. Seems useful...

```{r structure}
str(gapminder)
```

Very handy! Will keep this in mind for a conceptual data summary in future datasets.
Bear in mind that data.frames are preferred to arrays and matrices, essentially because they're more organised. 

Now we can unravel some interesting data frame info:

```{summary}
names(gapminder)
dim(gapminder) # shows dimensions i.e. row x col
ncol(gapminder) # number of columns i.e. variables
nrow(gapminder) # number of rows i.e. observations
summary(gapminder) # summary statistics
library(skimr)
skim(gapminder) # also does summary statistics, but let's compare...
```

So skimr seems to give a more thorough overview of the data. 

# dplyr basics

There's five (5) main functions to perform basic data manipulation. In every case, the first argument is a data frame, and the subsequent arguments describe what to do with the data frame. The result is a new data frame. 

```{r dplyr}
filter()    # pick observations by their values i.e. subsets row-wise
select()    # pick observations by their names
mutate()    # create new variables with functions of existing variables
summarise() # collapse many values down to a single summary
arrange()   # reorder the rows
group_by()  # changes scope of above functions from the entire dataset to group-by-group

```

First we'll practise filteringA:

```{r filter}
# Filter using a value
filter(gapminder, lifeExp < 29)

# Filter using a string
filter(gapminder, country == "Mexico")
```
But the == operator only works on one thing at a time, so if we want more than one country, we need to use the %in% operator. I think this is some kind of 'pipe'? Don't really remember though

```{r two}
filter(gapminder, country %in% c("Mexico", "Peru"))
```

Now a challenge - find the average life expectancy in Brazil between 1987 and 2007. 

```{r challenge}
#assign a variable
Brazil <- filter(gapminder, country == "Brazil", year > 1986)

#then find the mean of the new dataset
mean(Brazil$lifeExp)
```

Now we move onto the select() function, for columns.

```{r select}
select(gapminder, year, country, lifeExp)
select(gapminder, year, -continent, -lifeExp)
```

Then combine filter() and select() ...

```{r combine}

gap_cambodia <- filter(gapminder, cougntry == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)
```

# Now we're gonna learn about pipes!!

It removes the need to nest multi-operation commands inside each other, and you also don't need to make temporary variables. 
Ctrl + Shift + M
Think "and then"

```{r piping}
gapminder %>% head()

```
The outcome is the same as gapminder(head), but the syntax starts with the data and then the action, like a SVO sentence instead of VSO. 
So for comparison:

```{r cambodia comparison}

gap_cambodia <- filter(gapminder, cougntry == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

# becomes

gap_cambodia <- gapminder %>% filter(country == "Cambodia")
gap_cambodia2 <- gap_cambodia %>% select(-continent, -lifeExp)

#AND THEEEEEN to get rid of the temporary variables, we combine the operations:

gap_cambodia <- gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(-continent, -lifeExp)
```

So now this reads more intuitively: "Start with gapminder data, and then filter for Cambodia and then drop the variables continent and lifeExp."

Next we'll use mutate(), to add new variables. A good example would be to add an index or label column.

```{r mutate}
gapminder %>% 
  mutate(gdp = pop * gdpPercap)
```

Now, to practise:

```{r mutate new}
#Calculate the population in thousands for all Asian countries in 2007 and add it as a new column

gapminder %>% 
  filter(continent == "Asia", 
         year == 2007) %>% 
  mutate(pop_thousands = pop/1000) %>% 
  select(country, year, pop_thousands) #this line is optional, to tidy the dataset
  

```

Now we move to group_by(), a grouping variable.

```{r group}

gapminder %>% 
  filter(year == 2002) %>% 
  group_by(continent) %>% 
  mutate(cont_pop = sum(pop))

```

Using the summarise function will narrow the dataset to only columns that are included in the group_by()

```{r summarising a group}
gapminder %>% 
  group_by(continent) %>% 
  summarise(cont_pop = sum(pop)) %>% 
  ungroup()
```

We can also use it with more than one variable. (*I'm getting kind of confused here - bit of information overload so taking these notes to refer back to)

```{r more than one variable}

gapminder %>% 
  group_by(continent, year) %>% 
  summarise(cont_pop = sum(pop))

```

Next we can use the arrange() function to order columns by whichever variable we choose.

```{r arrange}

gapminder %>% 
  group_by(continent, year) %>% 
  summarise(cont_pop = sum(pop)) %>% 
  arrange(continent)

```

Now for an exercise, we will determine the maximmum GDP per continent across all years.

```{r gdp}

gapminder %>%
  mutate(gdp = pop * gdpPercap) %>% 
  group_by(continent) %>% 
  mutate(max_gdp = max(gdp)) %>% 
  filter(gdp == max_gdp)


```

Now we're walking through a new problem:

```{r Asia}
#Arrange df in descending order
asia_lifeExp <- gapminder %>% 
  filter(continent == "Asia") %>% 
  group_by(country) %>% 
  filter(lifeExp == max(lifeExp)) %>% 
  arrange(year)
```

Will need to practise this in the long term. 
